Appendix 5: example of index standardization of fishery-independent survey data with VAST

### 2021-12-15 doc/appendix-pcod-VAST-tweedie.Rmd
change rhoconfig to obtain conventional Tweedie model
Add to Tweedie appendix

RhoConfig <- c("Beta1" = 0, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
```

@@

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
pcod <- as.data.frame(pcod) # ensure not a tibble
fit <- fit_model(
  settings = settings,
  Lat_i = pcod[, "lat"],
  Lon_i = pcod[, "lon"],
  t_i = pcod[, "year"],
  b_i = pcod[, "density"],
  a_i = pcod[, "effort"],
  input_grid = input_grid,
  working_dir = paste0(here("doc", "appendix-VAST-tweedie"), "/")
)
```

@@

@@

```{r save-plots, echo = TRUE, message = FALSE, warning = FALSE}
plot(
  fit,
  check_residuals = FALSE,
  working_dir = paste0(here("doc", "appendix-VAST-tweedie"), "/")
)
```

@@

>>>>

RhoConfig <- c("Beta1" = 3, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
```

@@

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
pcod <- as.data.frame(pcod) # ensure not a tibble
f <- here("doc", "appendix-VAST-tweedie", "vast-cache.rds")
if (!file.exists(f)) {
  tictoc::tic()
  fit <- fit_model(
    settings = settings,
    Lat_i = pcod[, "lat"],
    Lon_i = pcod[, "lon"],
    t_i = pcod[, "year"],
    b_i = pcod[, "density"],
    a_i = pcod[, "effort"],
    input_grid = input_grid,
    working_dir = paste0(here("doc", "appendix-VAST-tweedie"), "/")
  )
  tictoc::toc()
  saveRDS(fit, file = f)
} else {
  fit <- readRDS(f)
}
```

@@

Fit sdmTMB model and compare:

```{r sdmTMB-example, echo = TRUE, cache=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
  mesh = fit$spatial_list$MeshList$isotropic_mesh)
tictoc::tic()
fit_sdmTMB <- sdmTMB(density ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh, family = tweedie(link = "log"),
  time = "year", spatiotemporal = "iid", spatial = "on")
tictoc::toc()

s1 <- fit$ParHat$beta1_ft[fit$ParHat$beta2_ft != 0]
s2 <- fit$ParHat$beta2_ft[fit$ParHat$beta2_ft != 0]
b_VAST <- as.numeric(s1 + s2)

b_sdmTMB <- tidy(fit_sdmTMB)

plot(b_sdmTMB$estimate, b_VAST);abline(0, 1)
cor(b_sdmTMB$estimate, b_VAST)

s <- fit$parameter_estimates$SD
vast_est1 <- as.list(s, "Estimate", report = FALSE)
vast_est2 <- as.list(s, "Estimate", report = TRUE)
vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
vast_sd2 <- as.list(s, "Std. Error", report = TRUE)

sdmtmb_est <- tidy(fit_sdmTMB, "ran_pars")

# range
sdmtmb_est$estimate[sdmtmb_est$term == "range"]
vast_est2$Range_raw2

# sigma_O
sdmtmb_est$estimate[sdmtmb_est$term == "sigma_O"]
vast_est1$L_omega2_z

# sigma_E
sdmtmb_est$estimate[sdmtmb_est$term == "sigma_E"]
vast_est1$L_epsilon2_z

# Tweedie p
sdmtmb_est$estimate[sdmtmb_est$term == "tweedie_p"]
1 + plogis(fit$ParHat$logSigmaM[1,1] )

# Tweedie dispersion
as.numeric(exp(fit$ParHat$beta1_ft))[1]
sdmtmb_est$estimate[sdmtmb_est$term == "phi"]
```

@@

```{r save-plots, echo = TRUE, message = FALSE, warning = FALSE}
if (!file.exists(here("doc", "appendix-VAST-tweedie", "Data_and_knots.png")) {
  plot(
    fit,
    check_residuals = FALSE,
    working_dir = paste0(here("doc", "appendix-VAST-tweedie"), "/")
  )
}
```

@@

We can compare the index we would get using sdmTMB. Since sdmTMB does not have built-in delta models, we need some way of combining the predictions across the two models. Here, we will do that by simulating from the joint parameter precision matrix and calculating the index with those draws. This should be approximately comparable with the bias-corrected index calculated above with VAST.

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
pred <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE, area = 4)
ind_sdmTMB <- get_index(pred)
```

Now, we can compare the indices. We will read in the VAST index from the `Index.csv` file created by the `plot.fit_model()` method above.

```{r index-compare, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
    se = Std..Error.for.ln.Estimate.) %>% 
  select(index, year, est, se) %>% 
  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
  mutate(upr = exp(log(est) + qnorm(0.975) * se))
sdm_i <- ind_sdmTMB %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(alpha = 0.3) +
  geom_line() + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE) +
  ylab("Biomass (t)")
```
